---
title: 'Perl 6 documentation repository through time: An initial report'
author: "JJ Merelo"
date: "26 de abril de 2018"
output: html_document
bibliography: perl6.bib
abstract: After getting a grant from the Perl Foundation for taking care of the repository where Perl 6 documents are hosted, I wanted to run an initial assessment of how this documentation is behaving in terms of community and evolution. One of the compromises I committed to was to engage with at least 2 issues a day, since the number of issues was initially around 300 and was increasing. So in this initial report I will take a look at these issues, who posts them, who solves them, how long does it take to close them. 
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggthemes)

issues <- read.csv("../data/issues.csv")
opens.gt.1 <- issues[issues$created_by %in%  names( table(issues$created_by)[table(issues$created_by) > 1]),]
closes.gt.1 <- issues[issues$closer %in%  names( table(issues$closer)[table(issues$closer) > 1]),]
issues$author_association <- as.factor(issues$author_association)
closes.gt.1 <- closes.gt.1[closes.gt.1$state == " closed",]
```

# Introduction

Perl 6  [@randal2004perl,@randal2004perl,@lenzperl] is a language whose development started earlier in this century and was released as a production ready system by Christmas 2015. It is an interpreted language with two parts, a virtual machine which supports and abstracts all basic functionalities in the language, and which so far includes an original virtual machine, MoarVM, as well as JVM, and a compiler for that virtual machine, currently called [Rakudo](https://rakudo.org). It is got a fast development cycle, with monthly releases.

For the purposes of this report, an interesting feature of Perl 6 is that its development is entirely voluntary-based. The Perl Foundation is the organizm in charge of supporting it, but there are very few, if any, full time developers working on it, or for that matter the documentation. That makes it an interesting case on free software development, all by itself.

Documentation is not, per se, part of the core of a language, however it is an essential feature to attract new users and keep existing users. Bad or non-existing documentation will substract users, or simply not attract them. Usually, however, it is not the first thing that pops into one's head when one want to devote resources to it. Not even volunteer developers. Perl 6, however, has considered it from the beginning a first-class citizen. The scarcity of resources and its cyclic nature has made the result uneven in frequency, if not in quality.

In this repor we will try to visualize the behavior of the community, and how the repository has behaved across time. 

# Who posts issues

One of the interesting things to see is who, effectively, posts issues. There are links to do that from every documentation page; however, are new users who find some problem encouraged to actually post it or are perl 6 developers mainly who do that?


```{r issuers,echo=FALSE}
ggplot(issues,aes(x=author_association,fill=state))+geom_bar() 
``` 
# References
