---
title: 'Perl 6 documentation repository through time: An initial report'
author: "JJ Merelo"
date: "29 de abril de 2018"
output:
  pdf_document: default
  html_document: default
bibliography: perl6.bib
abstract: After getting a grant from the Perl Foundation for taking care of the repository
  where Perl 6 documents are hosted, I wanted to run an initial assessment of how
  this documentation is behaving in terms of community and evolution. One of the compromises
  I committed to was to engage with at least 2 issues a day, since the number of issues
  was initially around 300 and was increasing. So in this initial report I will take
  a look at these issues, who posts them, who solves them, how long does it take to
  close them.
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggthemes)

issues <- read.csv("../data/issues.csv")
opens.gt.1 <- issues[issues$created_by %in%  names( table(issues$created_by)[table(issues$created_by) > 1]),]
closes.gt.1 <- issues[issues$closer %in%  names( table(issues$closer)[table(issues$closer) > 1]),]
issues$author_association <- as.factor(issues$author_association)
closes.gt.1 <- closes.gt.1[closes.gt.1$state == " closed",]
issues$created_at <- as.Date(issues$created_at)
issues$closed_at <- as.Date(issues$closed_at)
closed.issues <- issues[issues$state == ' closed',]
closed.issues$duration <- closed.issues$closed_at - closed.issues$created_at

```

# Introduction

Perl 6  [@randal2004perl,@tang2007perl,@lenzperl] is a language whose development started earlier in this century and was released as a production ready system by Christmas 2015. It is an interpreted language with two parts, a virtual machine which supports and abstracts all basic functionalities in the language, and which so far includes an original virtual machine, MoarVM, as well as JVM, and a compiler for that virtual machine, currently called [Rakudo](https://rakudo.org). It is got a fast development cycle, with monthly releases.

For the purposes of this report, an interesting feature of Perl 6 is that its development is entirely voluntary-based. The Perl Foundation is the organism in charge of supporting it, but there are very few, if any, full time developers working on it, or for that matter the documentation. That makes it an interesting case on free software development, all by itself.

Documentation is not, per se, part of the core of a language, however it is an essential feature to attract new users and keep existing users. Bad or non-existing documentation will subtract users, or simply not attract them. Usually, however, it is not the first thing that pops into one's head when one want to devote resources to it. Not even volunteer developers. Perl 6, however, has considered it from the beginning a first-class citizen. The scarcity of resources and its cyclic nature has made the result uneven in frequency, if not in quality.

In this report we will try to visualize the behavior of the community, and how the repository has behaved across time. 

# How many issues are there

What prompted me to apply for the grant was the number of open issues in the repository. There were close to 300, and the number seemed to be increasing. This is by no means unusual, even in managed projects. Besides, the nature of issues, which are never critical, invites to postpone them, so they tend to crop up to a certain number. The problem is that beyond a certain number of issues, duplicates are inevitable, and prioritization is complicated, with the latest issues becoming more attractive for fixing than anything else. But here is how the number of issues evolved in time, to the situation roughly around the time the grant was started.

```{r timeline,echo=FALSE}
issues.t <- read.csv("../data/issues-timeline.csv")

issues.t$Time <- as.Date(issues.t$Time)
ggplot(issues.t, aes(x=Time,y=Open.Issues))+geom_line()+theme_tufte()
``` 

We can observe an approximately stable situation until mid-2016, when the number of issues starts to climb until it reaches the current situation of around 260 open issues. Some dips in the number of issues match the [monthly bug squash day](https://github.com/rakudo/rakudo/wiki/Monthly-Bug-Squash-Day), which every two months is devoted to this repository. However important, these bug-squash days take the open issue situation one month back, approximately. 

This situation is obviously due to the fact that more issues are created than closed. Below is the number of issues created per month and its current situation.

```{r issuecreation,echo=FALSE,message=FALSE}
issues$creationMonth <- as.Date(cut(issues$created_at, breaks="month"))
ggplot(issues,aes(x=creationMonth,fill=state))+geom_bar(stat='count')
``` 

There is no clear trend here, although in general, several issues are open every day, with some months having more than 100 issues open. This roughly corresponds with usage of the language: as the language increases its popularity, so people go to the site and report issues. The fact that usage, or at least issue creation, is not matched by the same number of collaborators in the repository causes this imbalance and the number of issues to grow.

# Who posts issues

One of the interesting things to see is who, effectively, posts issues. There are links to do that from every documentation page; however, are new users who find some problem encouraged to actually post it or are Perl 6 developers mainly who do that?
Well try to find that out below, plotting the number of issues per kind of user and how they are doing. 

```{r issuers,echo=FALSE}
ggplot(issues,aes(x=author_association,fill=state))+geom_bar() 
``` 
At the same time, we also plot whether those issues are open or closed. First thing that stands out is that the `NONE` category is the smallest, with very few issues, most of them are closed. `NONE` includes users that have no other role in the repository, and it is quite clear from this that they are a slim minority. However, users with no association to the repository can become `CONTRIBUTOR`s once they create a pull request which is accepted; that is the second category, with close to 200 issues. Of course, the repository owner and members with write privileges constitute the vast majority. 
This probably means that most users who care enough to report issues are already inside the community of Perl 6 developers, or have eventually integrated into it. We are a very open community, so that might be a good thing. But that eventually talks more about the size of the community of developers integrated in the Perl 6 repos as compared with the total community of developers. It is a snapshot of a situation in the early stages of the release, anyhow. 
Let us look at the same chart in a different way. 
```{r staters,echo=FALSE}
ggplot(issues,aes(x=state,fill=author_association))+geom_bar(position="fill") 
``` 
What we see here are very slight differences in percentages of issues open or closed depending on the contributor. Basically it is telling us that `OWNER`-created issues tend to form a bigger percentage of open than closed issues; in other words, issues from owners are slightly more likely to stay open, at least if you look at it statically.

# Time to response.

Another interesting metric is how long issues tend to stay open. This number is expected to be high, since at any one particular moment there might be around one quarter of all issues created open, but the distribution of duration might tell us more about the dynamics of the repository.
```{r duration,echo=FALSE,message=FALSE}
duration.summary <- summary(as.numeric(closed.issues$duration))
closed.issues$duration = as.numeric(closed.issues$duration)
closed.issues$transformed = closed.issues$duration+1
ggplot(closed.issues,aes(x=transformed))+geom_histogram()+scale_x_log10()
``` 
This histogram shows that most issues are solved straight away, in the very first day. However, there are a few issues that take a long time. There is a small bump around one year and even a few that take more than 3 years. The median duration is `r duration.summary['Median']`; that is the number of days needed to fix 50% of the issues. The average is higher, `r duration.summary['Mean']`, due to the highly skewed distribution. And that highly skewed distribution is also a problem. 25% of the issues take longer than `r duration.summary['3rd Qu.']` days to be closed. That is why, also, at any given time, there is more or less around the same number lingering as not closed in the repository.
```{r duration2,echo=FALSE,message=FALSE}
ggplot(closed.issues,aes(x=transformed))+geom_point(stat='count')+scale_x_log10()+scale_y_log10()
``` 
This chart, that represents in abscissas the number of days and in the `y` axis the number of issues taking those number of days, shows a power law. The bigger the number of days, the less frequent it is, but it also exhibits a long tail, with issues with many days creating it. This *long tail* has already been observed in other software projects [@dalle2007different], and is explained by the existence of *superbugs* and different *fixing regimes*, but also to the fact that in many case fixing one issue leads to discovering other similar issues, be it in the same file or with similar nature, and to fix them too. The fact that in this case the repository is mostly text does not change the behavior, which is thus confirmed in the case of the Perl 6 documentation repository. It might be the case that the bugs that take longer have a different nature, but that will be the object of other report. 

# Conclusions

This is an initial data analysis of the dynamics in the [Perl 6 documentation repository](https://github.com/perl6/doc) and its situation until the beginning of April 2018. It is a completely volunteer-based repository, with most people helping in their spare time. Even so, it has got a high degree of completion of issues. 

Most repositories show some kind of emergent, power-law behavior [@merelo16self,@merelo17vue,@MereloCMVCF17]; this power law behavior, which had already been observed by other researchers, appears also in the time-to-solution of the issues in this repository. Long times to solution should be avoided, however, which would also shorten the list of open issues; this will arrive only via having some kind of permanent human resources assigned to this particular repository and make these resources increase with the popularity of the language. Right now the situation is more or less under control, with the number of issues increasing more slowly in the last months, but it could get out of hand very fast if there is a sudden surge in interest for the language. 

As future work, we will update this report with data during the first few weeks of the grant and after its final, to check if the curation grant awarded to the author has had any noticeable impact on these numbers and the trend they indicate. 

All data for this report, as well as the report itself, is [available from its repository](https://github.com/JJ/TPF-grant) under a free license. All data processing and extraction has been done in Perl 6, with statistics and charts in R. This report has been written using RMarkdown.

## Acknowledgements

This report has been supported by The Perl Foundation under [grant *Curating and improving Perl 6 documentation*](http://news.perlfoundation.org/2018/02/grant-proposal-curating-and-im.html). 

# References
